---
title: "<div style='text-align:right; font-size:14pt;'><a href='blog.html'>Back to Blog Home</a></div>Analysis of County Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rvest)
library(dplyr)
library(magrittr)
library(readr)
library(stringr)
library(ggplot2)
```

Recently in an advanced methods course my professor was discussing data on some of the most populous counties in the US.
She then asked how many counties we thought there were in California (where I am located, and the most populous US state).
My initial guess was somewhere between 20 and 30, but a quick google search showed us that there are actually **58** counties in California.
This seemed like a lot, but when I researched it further, I found that many other states have much more than California, despite being less populous.
In this post, we will webscrape some data from Wikipedia and perform a brief analysis on it in R.

## Getting the data

I found a great table on the [Wikipedia page located here](https://en.wikipedia.org/wiki/County_(United_States)), under the "Statistics" section.
We will use `{rvest}` and `{xml2}` to pull the data into R.

First, let's get our libraries:
```{r, eval=FALSE}
library(rvest)
library(dplyr)
library(magrittr)
library(readr)
library(stringr)
library(ggplot2)
```

And then we can read in the data.
```{r}
url <- "https://en.wikipedia.org/wiki/County_(United_States)"

county_raw <- read_html(url) %>%
  html_node("table.wikitable.sortable") %>%
  html_table()
head(county_raw)
```

Now that we have the table in R, we need to clean the data.
From the head, we can see that the original column titles got split into the table header and the first row of the data.
To fix this, we will remove the first row and then rename all columns.
Furthermore, we will convert columns to the correct type and make all area measurements in square miles.

```{r}
county_clean <- county_raw %>%
  set_colnames(
    c(
      "state",
      "population_2019",
      "land_area",
      "counties",
      "equivalents",
      "total_subdivisions",
      "avg_county_population",
      "avg_county_size"
    )
  ) %>%
  slice(-1) %>%
  mutate(
    population_2019 = parse_number(population_2019),
    counties = parse_number(counties),
    equivalents = parse_number(equivalents),
    total_subdivisions = parse_number(total_subdivisions),
    avg_county_population = parse_number(avg_county_population),
    land_area = str_extract(land_area, "^.* ") %>% parse_number(),
    avg_county_size = str_extract(avg_county_size, "^.* ") %>% parse_number()
  ) %>%
  suppressWarnings() %>% # known NA's caused by - in numeric columns
  slice(1:51) # just get the 50 states and DC, no territories or totals
  
head(county_clean)
```

That looks much better and way more useable. Hooray!

## Exploring the data

We will explore two main relationships here, the relationship between state population and the number of counties, and the relationship between state land area and the number of counties.
Plots of these two relationships are shown below.

```{r}
county_clean %>%
  ggplot() +
  aes(x = population_2019, y = total_subdivisions) +
  geom_point() +
  xlab("State Population") +
  ylab("# of Counties (or Equivalents)") +
  scale_x_continuous(
    labels = function(x) {
      str_c(x / 1000000, " Milliion")
    }
  ) +
  theme_bw()
```

```{r}
county_clean %>%
  ggplot() +
  aes(x = land_area, y = total_subdivisions) +
  geom_point() +
  xlab("State Land Area (Thousands of Square Miles)") +
  ylab("# of Counties (or Equivalents)") +
  scale_x_continuous(
    labels = function(x) {
      str_c(x / 1000)
    }
  ) +
  theme_bw()
```

We can also look at the correlation between these variables:
```{r}
cor(county_clean$total_subdivisions, county_clean$population_2019)
cor(county_clean$total_subdivisions, county_clean$land_area)
```

And maybe even add a regression line to the above plots, although it should be noted that the regression assumptions are definitely not met for either of these relationships.
Further analysis would be needed to create a valid model of the relationship.

```{r}
county_clean %>%
  ggplot() +
  aes(x = population_2019, y = total_subdivisions) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("State Population") +
  ylab("# of Counties (or Equivalents)") +
  scale_x_continuous(
    labels = function(x) {
      str_c(x / 1000000, " Milliion")
    }
  ) +
  theme_bw()
summary(lm(total_subdivisions ~ population_2019, data = county_clean))

county_clean %>%
  ggplot() +
  aes(x = land_area, y = total_subdivisions) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("State Land Area (Thousands of Square Miles)") +
  ylab("# of Counties (or Equivalents)") +
  scale_x_continuous(
    labels = function(x) {
      str_c(x / 1000)
    }
  ) +
  theme_bw()
summary(lm(total_subdivisions ~ land_area, data = county_clean))
```

## Conclusion

There seems to be a lot that could be done with this data, but here we just showed how to get the data, and performed a brief analysis on it.
Hope you found it helpful and interesting!