---
title: "<div style='text-align:right; font-size:14pt;'><a href='portfolio.html'>Back to Portfolio Home</a></div>Presentation Portfolio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
```

```{css}
h2, h3, h4, h5, h6{
  text-align:center;
}
```

<hr style="border-top: 1px solid black">

## mSHAP Presentation at IME 2021 {#mshap}

```{r}
fluidRow(
  column(
    width = 6,
    offset = 3,
    tags$a(
      tags$img(src = "resources/portfolio/mshap_pres.PNG", width = "100%", style = "padding-bottom: 10px;"),
      href = "resources/mshap_slides.pdf"
    )
  )
)
```


The [mSHAP research](https://srmatth.github.io/research-mshap.html) was presented at the 24th International Insurance: Mathematics and Economics Conference on July 8th, 2021. [View the slides here](resources/mshap_slides.pdf) or [see a transcript of the talk here.](resources/mshap_talk.pdf)
The actual talk can be viewed (with an actuview account) [at this website](https://www.actuview.com/video/mSHAP-Explainability-of-Two-Part-Models/a70f5b18d1f1e39cd51d65f73440b6a1).

<hr style="border-top: 1px solid black">

## Beach Volleyball Clustering Analysis {#volleyball}

```{r}
fluidRow(
  column(
    width = 6,
    offset = 3,
    tags$a(
      tags$img(src = "resources/portfolio/volleyball_partners.jpg", width = "100%", style = "padding-bottom: 10px;"),
      href = "resources/volleyball_partners.pdf"
    )
  )
)
```


A small personal project I created to submit a poster to the [UConn Sports Analytics Symposium 2021 (UCSAS)](https://statds.org/events/ucsas2021/index.html).
The poster documents the clustering of men's beach volleyball players into distinct groups and then analyzing the performance of different pairings of the groups.
Because of the virtual format of the conference, "posters" consisted of four slides.
[See the poster here.](resources/volleyball_partners.pdf)

<hr style="border-top: 1px solid black">


