---
title: "<div style='text-align:right; font-size:14pt;'><a href='portfolio.html'>Back to Portfolio Home</a></div>Data Visualization Portfolio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
```

```{css}
h2, h3, h4, h5, h6{
  text-align:center;
}
```

This page contains a curated portfolio of data visualizations that I have created.

<hr style="border-top: 1px solid black">

## Tidy Tuesday Data Visualizations

Every week, [rfordatascience](https://github.com/rfordatascience/tidytuesday) releases a new data set specifically geared for data visualization.
I greatly enjoy giving myself an hour to see what I can come up with from the data, and I'll display some of my best results here, with the most recent ones at the top.

### Lemurs (2021-08-24) {#lemur}

```{r}
fluidRow(
  style = "padding: 5px; background-color:black;",
    a(
      tags$img(src = "resources/portfolio/2021-08-24.jpeg", width = "100%"),
      href = "resources/portfolio/2021-08-24.jpeg"
    )
)
```

<br/>

### Olympic Medalists (2021-07-27) {#olympic}

```{r}
fluidRow(
  style = "padding: 5px; background-color:black;",
    a(
      tags$img(src = "resources/portfolio/2021-07-27.jpg", width = "100%"),
      href = "resources/portfolio/2021-07-27.jpg"
    )
)
```

<hr style="border-top: 1px solid black">

